On the specific recursive structure of iterated parametricity

### 2025-10-05 paper/paper.tex
paper: alternate, simpler explanation

\begin{document}
\title{On the specific recursive structure of iterated parametricity}
\begin{authgrp}
  \author{Hugo Herbelin}
  \affiliation{Université Paris Cité, Inria, CNRS, IRIF, Paris
    \email{Hugo.Herbelin@inria.fr}}
  \author{Ramkumar Ramachandra}
  \affiliation{Unaffiliated \email{r@artagnon.com}}
\end{authgrp}
\begin{abstract}
  In a previous paper, we gave equations describing the construction of semi-simplicial and semi-cubical sets via iterated parametricity. The construction was formalised in the Rocq (formerly known as Coq) proof assistant building $n$-truncated sets using an induction on a large structure embedding equational reasoning.

@@

\renewcommandx{\framep}[2][1,2]{\prim{frame}[][#1][#2][][]}
\renewcommandx{\layer}[2][1,2]{\prim{layer}[][#1][#2][][]}
\renewcommandx{\painting}[2][1,2]{\prim{painting}[][#1][#2][][]}
\renewcommandx{\restrf}[4][1,2,3,4]{\restr{frame}[][#3][#4][#1][#2][][]}
\renewcommandx{\restrl}[4][1,2,3,4]{\restr{layer}[][#3][#4][#1][#2][][]}
\renewcommandx{\restrp}[4][1,2,3,4]{\restr{painting}[][#3][#4][#1][#2][][]}

\section{The recursive structure of iterated parametricity}
Iterated parametricity (in indexed form) produces a family of type
families of the following form:
\begin{equation*}
  \begin{array}{llr}
    X_0 & :              & \U                                                                        \\
    X_1 & :              & X_0 \times X_0 \rightarrow  \U                                            \\
    X_2 & : \Pi a b c d. & X_1(a,b) \times X_1 (c,d) \times X_1(a,c) \times X_1 (b,d) \rightarrow \U \\
    \ldots
  \end{array}
\end{equation*}

Such family can be recursively described by the equations given on Table ?.

>>>>

\begin{document}
\title{On the specific recursive structure of iterated parametricity}
\begin{authgrp}
  \author{Hugo Herbelin}
  \affiliation{Université Paris Cité, Inria, CNRS, IRIF, Paris
    \email{Hugo.Herbelin@inria.fr}}
  \author{Ramkumar Ramachandra}
  \affiliation{Unaffiliated \email{r@artagnon.com}}
\end{authgrp}
\begin{abstract}
  In a previous paper, we gave equations describing the construction of semi-simplicial and semi-cubical sets via iterated parametricity. The construction was formalised in the Rocq (formerly, Coq) proof assistant building $n$-truncated sets using an induction on a large structure embedding equational reasoning.

@@

\renewcommandx{\framep}[2][1,2]{\prim{frame}[][#1][#2][][]}
\renewcommandx{\layer}[2][1,2]{\prim{layer}[][#1][#2][][]}
\renewcommandx{\painting}[2][1,2]{\prim{painting}[][#1][#2][][]}
\renewcommandx{\restrf}[4][1,2,3,4]{\restr{frame}[][#3][#4][#1][#2][][]}
\renewcommandx{\restrl}[4][1,2,3,4]{\restr{layer}[][#3][#4][#1][#2][][]}
\renewcommandx{\restrp}[4][1,2,3,4]{\restr{painting}[][#3][#4][#1][#4][][]}
\renewcommandx{\cohf}[6][1,2,3,4,5,6]{\coh{frame}[][#3][#4][#5][#6][#1,#2][][]}
\renewcommandx{\cohl}[6][1,2,3,4,5,6]{\coh{layer}[][#3][#4][#5][#6][#1,#2][][]}
\renewcommandx{\cohp}[6][1,2,3,4,5,6]{\coh{painting}[][#3][#4][#5][#6][#1,#2][][]}

In \citep{hr25}, the well-foundedness was justified by induction on $n$ and subinductions, one on $p$ to mutually build $\framep[n][p]$, $\restrf[n][p]$ and $\cohf[n][p]$, and three others to build each of $\painting[n][p]$, $\restrp[n][p]$ and $\cohp[n][p]$ by induction on $n-p$.

Since $\restrf[n][p]$ is referring to $\framep[n][p]$ and $\framep[n-1][p]$, and $\cohf[n][p]$ is referring to $\restrf[n][p]$ and $\restrf[n-1][p]$, thus also to $\framep[n][p]$, $\framep[n-1][p]$ and $\framep[n-2][p]$, we built all of these 6 components at the same time. But equations were also needed to state the properties of $\framep[n][p]$, $\framep[n-1][p]$ and $\restrf[n][p]$ so that we were building by induction on $n+p$ (that is on $n$, using the convention of \citep{hr25}) the following block of information:

\begin{align*}
  \framep[n][p]      & : \U                                                                                                                 \\
  \framep[n-1][p]    & :  \U                                                                                                                \\
  \framep[n-2][p]    & : \U                                                                                                                 \\
  \restrf[n][p]      & : \Pi q.\, \Pi \omega.\, \framep[n][p] \rightarrow \framep[n-1][p]                                                   \\
  \restrf[n-1][p]    & : \Pi q.\, \Pi \omega.\, \framep[n-1][p] \rightarrow \framep[n-2][p]                                                 \\                                                                 \\
  \cohf[n][p]        & : \Pi qr\epsilon\omega.\, \restrf[n-1][p][\epsilon][q] \circ \restrf[n][p][\omega][r] =                              \\
                     & \;\;\;\;\restrf[n-1][p][\omega][r] \circ \restrf[n][p][\epsilon][q+1]                                                \\
  \mathsf{hypF0}     & : \framep[n][0]         = \unitpoint
  \\
  \mathsf{hypFS0}    & : \framep[n][p+1]       = \Sigma d: \framep[n+1][p]. \Pi \omega.\, \painting[n][p] (\restrf[n+1][p](d))              \\
  \mathsf{hypPrevF}  & : \framep[n-1][0]                                                                                       = \unitpoint \\
  \mathsf{hypPrevFS} & : \framep[n-1][p+1] = \Sigma d: \framep[n][p]. \Pi \omega.\, \painting[n-1][p] (\restrf[n][p](d))                    \\
  \mathsf{hypRF0}    & : \restrf[n][0]        = \ldots \texttt{depending on hypF0}                                                          \\
  \mathsf{hypRFS}    & : \restrf[n][p+1]      = \ldots \texttt{depending on hypFS and hypPrevFS}                                            \\
  \ldots             & \;\;\;\;\texttt{fields for painting, restrf, cohp}
\end{align*}

In the current paper and associated formalisation, we replace the equational properties by definitions, building instead by induction a block of the form:

\begin{align*}
  \framep[n-2][p] & : \U                                                                                    \\
  \restrf[n-1][p] & : \Pi q \leq n+p.\, \framep[n-1][p] \rightarrow \framep[n-2][p]                         \\
  \cohf[n][p]     & : \Pi qr\epsilon\omega.\, \restrf[n-1][p][\epsilon][q] \circ \restrf[n][p][\omega][r] = \\
                  & \;\;\;\;\restrf[n-1][p][\omega][r] \circ \restrf[n][p][\epsilon][q+1]                   \\
  \ldots          & \;\;\;\;\texttt{fields for painting, restrf, cohp}
\end{align*}
where $\framep[n][p]$, $\framep[n-1][p]$ and $\restrf[n][p]$ are now appropriate definitions.

Turning $\framep[n][p]$, $\framep[n-1][p]$ and $\restrf[n][p]$ into definitions is however tricky: the definition of $\framep[n-1][p]$, which is by induction on $p$ (for all $n$), depends on $\restrf[n-1][p]$ whose type depends on $\framep[n][p-1]$. This means that $\framep[n-1][p]$ and the type of $\restrf[n-1][p]$ have to be mutually defined.
